---
title: "tidy.js for R"
output: js4shiny::html_document_js
---

```{=html}
<a href="https://github.com/gadenbuie/tidyjs-r" class="github-corner" aria-label="View source on GitHub"><svg width="80" height="80" viewBox="0 0 250 250" style="fill:#eaecee; color:#2f4858; position: fixed; top: 0; border: 0; right: 0;" aria-hidden="true"><path d="M0,0 L115,115 L130,115 L142,142 L250,250 L250,0 Z"></path><path d="M128.3,109.0 C113.8,99.7 119.0,89.6 119.0,89.6 C122.0,82.7 120.5,78.6 120.5,78.6 C119.2,72.0 123.4,76.3 123.4,76.3 C127.3,80.9 125.5,87.3 125.5,87.3 C122.9,97.6 130.6,101.9 134.4,103.2" fill="currentColor" style="transform-origin: 130px 106px;" class="octo-arm"></path><path d="M115.0,115.0 C114.9,115.1 118.7,116.5 119.8,115.4 L133.7,101.6 C136.9,99.2 139.9,98.4 142.2,98.6 C133.8,88.0 127.5,74.4 143.8,58.0 C148.5,53.4 154.0,51.2 159.7,51.0 C160.3,49.4 163.2,43.6 171.4,40.1 C171.4,40.1 176.1,42.5 178.8,56.2 C183.1,58.6 187.2,61.8 190.9,65.4 C194.5,69.0 197.7,73.2 200.1,77.6 C213.8,80.2 216.3,84.9 216.3,84.9 C212.7,93.1 206.9,96.0 205.4,96.6 C205.1,102.4 203.0,107.8 198.3,112.5 C181.9,128.9 168.3,122.5 157.7,114.1 C157.9,116.9 156.7,120.9 152.7,124.9 L141.0,136.5 C139.8,137.7 141.6,141.9 141.8,141.8 Z" fill="currentColor" class="octo-body"></path></svg></a><style>.github-corner:hover .octo-arm{animation:octocat-wave 560ms ease-in-out}.github-corner:hover .octo-body,.github-corner:hover .octo-arm{color: #993d70; transition: color 500ms ease-in-out;}@keyframes octocat-wave{0%,100%{transform:rotate(0)}20%,60%{transform:rotate(-25deg)}40%,80%{transform:rotate(10deg)}}@media (max-width:500px){.github-corner:hover .octo-arm{animation:none}.github-corner .octo-arm{animation:octocat-wave 560ms ease-in-out}}</style>
```

[use_tidyjs()]: https://github.com/gadenbuie/tidyjs-r/blob/main/R/tidyjs.R

```{r setup, include=FALSE}
tidyjs::use_tidyjs()
xaringanExtra::use_clipboard(selector = "pre.r,pre.markdown,pre.js")
```

Install tidyjs from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("gadenbuie/tidyjs-r")
```

```{r child="../man/fragments/tidyjs-usage.Rmd"}
```

```{js, js_redirect = FALSE, echo = FALSE}
const { tidy, mutate, arrange, desc, summarize, sum, groupBy } = Tidy
```

The examples below are actually running in your browser.
The results of each JavaScript chunk, if executed, are shown below the JavaScript chunk with a gray left border.

```{js}
console.log('Ready to tidy?')
```

---

## [tidy.js](https://github.com/pbeshai/tidy)

[![CircleCI](https://img.shields.io/circleci/build/gh/pbeshai/tidy)](https://app.circleci.com/pipelines/github/pbeshai/tidy)
[![npm](https://img.shields.io/npm/v/@tidyjs/tidy)](https://www.npmjs.com/package/@tidyjs/tidy)

**Tidy up your data with JavaScript!** Inspired by [dplyr](https://dplyr.tidyverse.org/) and the [tidyverse](https://www.tidyverse.org/), tidy.js attempts to bring the ergonomics of data manipulation from R to javascript (and typescript). The primary goals of the project are:

* **Readable code**. Tidy.js prioritizes making your data transformations readable, so future you and your teammates can get up and running quickly.

* **Standard transformation verbs**. Tidy.js is built using battle-tested verbs from the R community that can handle any data wrangling need.

* **Work with plain JS objects**. No wrapper classes needed — all tidy.js needs is an array of plain old-fashioned JS objects to get started. Simple in, simple out.

Secondarily, this project aims to provide acceptable types for the functions provided.


#### Quick Links

* [GitHub repo](https://github.com/pbeshai/tidy)
* [Project homepage](https://pbeshai.github.io/tidy)
* [API reference documentation](https://pbeshai.github.io/tidy/docs/api/tidy)
* [Playground](https://pbeshai.github.io/tidy/playground)
* [Observable Notebook](https://observablehq.com/@pbeshai/tidy-js-intro-demo)
* [GitHub Discussions for Q&A](https://github.com/pbeshai/tidy/discussions)
* [CodeSandbox showing basic HTML usage (UMD)](https://codesandbox.io/s/tidyjs-umd-example-n1g4r?file=/index.html)

#### Related work

Be sure to check out a very similar project, [Arquero](https://github.com/uwdata/arquero), from [UW Data](https://idl.cs.washington.edu/). 


## Getting started

```{js}
const { tidy, mutate, arrange, desc } = Tidy

window.data = [
  { a: 1, b: 10 }, 
  { a: 3, b: 12 }, 
  { a: 2, b: 10 }
]

const results = tidy(
  data, 
  mutate({ ab: d => d.a * d.b }),
  arrange(desc('ab'))
)

results
```

All tidy.js code is wrapped in a **tidy flow** via the `tidy()` function. The first argument is the array of data, followed by the transformation verbs to run on the data. The actual functions passed to `tidy()` can be anything so long as they fit the form:

```
(items: object[]) => object[]
```

For example, the following is valid:

```{js}
tidy(
  data, 
  items => items.filter((d, i) => i % 2 === 0),
  arrange(desc('value'))
)
```

All tidy verbs fit this style, with the exception of exports from groupBy, discussed below.

### Grouping data with groupBy

Besides manipulating flat lists of data, tidy provides facilities for wrangling grouped data via the `groupBy()` function.

```{js}
const { tidy, summarize, sum, groupBy } = Tidy

const data = [
  { key: 'group1', value: 10 }, 
  { key: 'group2', value: 9 }, 
  { key: 'group1', value: 7 }
]

tidy(
  data,
  groupBy('key', [
    summarize({ total: sum('value') })
  ])
)
```

The output is:
```js
[
  { "key": "group1", "total": 17 },
  { "key": "group2", "total": 9 },
]
```

The `groupBy()` function works similarly to `tidy()` in that it takes a flow of functions as its second argument (wrapped in an array). Things get really fun when you use groupBy's *third* argument for exporting the grouped data into different shapes. 

For example, exporting data as a nested object, we can use `groupBy.object()` as the third argument to `groupBy()`.
 
```{js}
window.data = [
  { g: 'a', h: 'x', value: 5 },
  { g: 'a', h: 'y', value: 15 },
  { g: 'b', h: 'x', value: 10 },
  { g: 'b', h: 'x', value: 20 },
  { g: 'b', h: 'y', value: 30 },
]

tidy(
  data,
  groupBy(
    ['g', 'h'], 
    [
      mutate({ key: d => `${d.g}${d.h}`})
    ], 
    groupBy.object() // <-- specify the export
  )
);

```

The manually formatted output is:

```js
{
  "a": {
    "x": [{"g": "a", "h": "x", "value": 5, "key": "ax"}],
    "y": [{"g": "a", "h": "y", "value": 15, "key": "ay"}]
  },
  "b": {
    "x": [
      {"g": "b", "h": "x", "value": 10, "key": "bx"},
      {"g": "b", "h": "x", "value": 20, "key": "bx"}
    ],
    "y": [{"g": "b", "h": "y", "value": 30, "key": "by"}]
  }
}
```

Or alternatively as `{ key, values }` entries-objects  via `groupBy.entriesObject()`:

```{js}
tidy(data,
  groupBy(
    ['g', 'h'], 
    [
      mutate({ key: d => `${d.g}${d.h}`})
    ], 
    groupBy.entriesObject() // <-- specify the export
  )
);
```

The manually formatted output is:

```js
[
  {
    "key": "a",
    "values": [
      {"key": "x", "values": [{"g": "a", "h": "x", "value": 5, "key": "ax"}]},
      {"key": "y", "values": [{"g": "a", "h": "y", "value": 15, "key": "ay"}]}
    ]
  },
  {
    "key": "b",
    "values": [
      {
        "key": "x",
        "values": [
          {"g": "b", "h": "x", "value": 10, "key": "bx"},
          {"g": "b", "h": "x", "value": 20, "key": "bx"}
        ]
      },
      {"key": "y", "values": [{"g": "b", "h": "y", "value": 30, "key": "by"}]}
    ]
  }
]
```

It's common to be left with a single leaf in a groupBy set, especially after running summarize(). To prevent your exported data having its values wrapped in an array, you can pass the `single` option to it.

```{js}
tidy(data,
  groupBy(['g', 'h'], [
    summarize({ total: sum('value') })
  ], groupBy.object({ single: true }))
);
```

The manually formatted output is:

```js
{
  "a": {
    "x": {"total": 5, "g": "a", "h": "x"},
    "y": {"total": 15, "g": "a", "h": "y"}
  },
  "b": {
    "x": {"total": 30, "g": "b", "h": "x"},
    "y": {"total": 30, "g": "b", "h": "y"}
  }
}
```

Visit the [API reference docs](https://pbeshai.github.io/tidy/docs/api/tidy) to learn more about how each function works and all the options they take. Be sure to check out the `levels` export, which can let you mix-and-match different export types based on the depth of the data. For quick reference, other available groupBy exports include: 

* groupBy.entries()
* groupBy.entriesObject()
* groupBy.grouped()
* groupBy.levels()
* groupBy.object()
* groupBy.keys()
* groupBy.map()
* groupBy.values()

---


#### Shout out to Netflix

I want to give a big shout out to [Netflix](https://research.netflix.com/), my current employer, for giving me the opportunity to work on this project and to open source it. It's a great place to work and if you enjoy tinkering with data-related things, I'd strongly recommend checking out [our analytics department](https://research.netflix.com/research-area/analytics).
– [Peter Beshai](https://peterbeshai.com/)
